Real-space computation of $E$/$B$-mode maps I: Formalism, Compact Kernels, and Polarized Filaments

### 2018-06-29 articles/formalism/article.tex
updated greens kernel definition, reordered fig 7, cleaned up revisits, cleaned up discussions about telescoping behaviour
small edits to introduction and filament section.

\begin{document}
%\title{Quantifying the locality of operators that translate Stokes $Q$ \& $U$ to scalars $E$ \& $B$}
%\title{Real space operators for Stokes $Q$ \& $U$ to scalars $E$ \& $B$ translation}
%\title{Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels}
\title{Real-space computation of $E$/$B$-mode maps I: \revisit{Formalism, Compact Kernels, and Polarized Filaments}}

\author[]{Aditya Rotti and}
\author[]{Kevin Huffenberger}
\affiliation[]{Department of Physics, Florida State University, Keen Physics Building, 77 Chieftan Way, Tallahassee, Florida, U.S.A.}
\emailAdd{adityarotti@gmail.com}
\emailAdd{khuffenberger@fsu.edu}

\abstract{
\noindent 
\revisit{We derive full-sky, real-space operators that convert between polarization Stokes $Q/U$ parameters to the coordinate-independent scalar $E$/$B$ modes that are widely used in Cosmic Microwave Background(CMB) and cosmic shear analysis. We also derive real space operators that decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations.  For all these real space operators we show that the kernels split naturally into angular and radial parts and we show explicitly how the radial extent of these kernels depends on the targeted band-limit. We show that the kernels can be interpreted either as a complex convolving beam or the Green's function when they are expressed as functions of the forward or inverse rotation Euler angles respectively. We show that an arbitrary radial dependence, satisfying the constraint that it vanishes at angular separations of $0$ and $\pi$, can produce $E/B$-like maps and that these are simply filtered versions of the standard $E/B$ maps.  This allows us to compute $E/B$ maps in real space with a compactly-supported kernel, an approach that can guarantee the avoidance of known foreground regions and can be employed in a massively parallel scheme at high-resolution. We show that the spin raising $\eth^2$ and lowering $\bar{\eth}^2$ operators are a special case of these generalized operators and present their band limited versions. We can recover the standard power spectrum of the polarized CMB sky by correcting the power spectra of these maps with a simple window function. We present a scheme to derive this window function for any radial dependence. The spatial structure of the real space operator provides great intuition for the $E/B$ structure of polarized, filamentary galactic foregrounds and we predict a non-zero $B$-mode signature that is expected from finite filaments in the sky. This paper is the first part in a series of papers that explore real-space computation of polarization modes and their applications.
  } 
  \\
  \\
%  We derive fully sky real space operators to translate Stokes Q \& U parameters to scalars $E$ \& $B$ and vice versa. We explicitly show that these local real space operator are fully characterized by the spin-0 $Y_{\ell 2}$ spherical harmonic functions. These real space operators make transparent the association of radial and tangential patterns of polarization with the E modes and that of  the clockwise and anti-clockwise pinwheel patterns of polarization with B-modes. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations. Using this new notation also allows us to derive real space operators which decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We use these analytical derivations us to quantify the non-local nature of the relation between the two representations of the polarization field. We present a prescription for generalizing these operators, which allow the non-local behavior of these operators to be a tunable parameter. We derive constraints on these generalized operators to ensure reliable recovery of the standard $E$ and $B$ fields and the corresponding power spectra. \revisit{This is part 1 of a series of papers where we have derived the real space operators. In the following papers we will explore some of their applications.} \comment{Needs better blending.}
}

>>>>

\begin{document}
%\title{Quantifying the locality of operators that translate Stokes $Q$ \& $U$ to scalars $E$ \& $B$}
%\title{Real space operators for Stokes $Q$ \& $U$ to scalars $E$ \& $B$ translation}
%\title{Real-space computation of $E$/$B$-mode maps I: Formalism and Compact Kernels}
\title{Real-space computation of $E$/$B$-mode maps I: Formalism, Compact Kernels, and Polarized Filaments}

\author[]{Aditya Rotti and}
\author[]{Kevin Huffenberger}
\affiliation[]{Department of Physics, Florida State University, Keen Physics Building, Tallahassee, Florida 32306}
\emailAdd{adityarotti@gmail.com}
\emailAdd{khuffenberger@fsu.edu}

\abstract{
\noindent 
    {We derive full-sky, real-space operators that convert between polarization Stokes $Q/U$ parameters and the coordinate-independent scalar $E$/$B$ modes that are widely used in Cosmic Microwave Background (CMB) and cosmic shear analysis. We also derive real space operators that decompose the measured Stokes parameters into those corresponding to $E$-modes and $B$-modes respectively, without ever evaluating the scalar fields themselves. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidates these derivations.  For all these real space operators we show that the kernels split naturally into angular and radial parts and we show explicitly how the radial extent of these kernels depends on the targeted band-limit. We show that the kernels can be interpreted either as a complex convolving beam or a Green's function when they are expressed as functions of the forward or inverse rotation Euler angles respectively. We show that an arbitrary radial function can produce $E/B$-like maps, provides it vanishes at the origin and the  antipodal point.  These maps are simply filtered versions of the standard $E/B$ maps.  We present a scheme to derive this window function for any radial dependence.  For this reason we can compute $E/B$ maps in real space with a compactly-supported kernel, an approach that can guarantee the avoidance of known foreground regions and could be employed in a massively-parallel scheme at high-resolution.  We show that the spin raising $\eth^2$ and lowering $\bar{\eth}^2$ operators are another special case of these generalized radial functions, and present their band limited versions. We can recover the standard power spectrum of the polarized CMB sky by correcting the power spectra of these maps with a simple window function. The spatial structure of the real space operators provides great intuition for the $E/B$ structure of polarized, filamentary galactic foregrounds.  We predict a non-zero $B$-mode signature that is expected from finite filaments in the sky. This paper is the first part in a series of papers that explore real-space computation of polarization modes and their applications.
  } 
  \\
  \\
%  We derive fully sky real space operators to translate Stokes Q \& U parameters to scalars $E$ \& $B$ and vice versa. We explicitly show that these local real space operator are fully characterized by the spin-0 $Y_{\ell 2}$ spherical harmonic functions. These real space operators make transparent the association of radial and tangential patterns of polarization with the E modes and that of  the clockwise and anti-clockwise pinwheel patterns of polarization with B-modes. We cast the standard CMB polarization analysis operators in a matrix-vector notation which elucidate these derivations. Using this new notation also allows us to derive real space operators which decompose the measured Stokes parameters into those corresponding to E-modes and B-modes respectively, without ever evaluating the scalar fields themselves. We use these analytical derivations us to quantify the non-local nature of the relation between the two representations of the polarization field. We present a prescription for generalizing these operators, which allow the non-local behavior of these operators to be a tunable parameter. We derive constraints on these generalized operators to ensure reliable recovery of the standard $E$ and $B$ fields and the corresponding power spectra. \revisit{This is part 1 of a series of papers where we have derived the real space operators. In the following papers we will explore some of their applications.} \comment{Needs better blending.}
}

### 2018-06-29 articles/formalism/sections/kernel_paintbrush.tex
small edits to introduction and filament section.

The $B$ part of the Green's function, traced along the filament, cancels itself except at the filament ends.  This results in a non-zero $B$ pattern for the filament.  For a North--South filament, the $B$-mode pattern is positive on the North-East and South-West, and negative in the North-West and South-East.  The size of this $B$-mode pattern is set by the dimensions of the filament, chiefly the width.

The non-zero $B$ result is somewhat surprising given that the polarization pattern is symmetric to both horizontal and vertical reflections through the filament center.  However, unlike a circular ring, this filament is not a configuration with a definite parity.  Since the scalar description of polarization is coordinate independent, the $E/B$ patterns do not depend on the orientation of the filament.  A filament inclined at $45^\circ$ will have a similarly inclined $E/B$ pattern, but different reflection symmetries.

Changing the polarization direction within the filament changes the $E/B$ patterns.  A $90^\circ$ rotation of the polarization with respect to the filament changes the sign of both $E$ and $B$.  In a polarization pattern aligned at $45^\circ$ to the filament, the $E$ pattern will swap with the $B$ pattern.  The work in \cite{Zaldarriaga2001a} correctly argued that linear filaments (infinite and without ends) can only produce $E$-power, but we find that for realistic filaments, the ends produce $B$-mode power with a clear signature.  Careful study of the $E/B$ mode power in filaments can provide insights into the orientation of the magnetic field with respect to the length of the filament.

The intuition from the real-space kernels holds also when we distort the shape of the filament.  If the filament were bent around into a circle, the positive and negative parts of the $B$ pattern will cancel, and we are left with a hoop of pure $E$ pattern. Here it is important to note that this cancellation will not happen for an ellipse or any loop of non-contant radius of curvature.  The same general description holds for a spiral-shaped filament, which can can be viewed as distortion of the straight filament.  The filament is highlighted by positive $E>0$.  The $E$-pattern is more negative on the interior of a curve than on the exterior, and the concentric rings of filamentary structure make an increasingly negative $E$ value inside. While the $B$-pattern is again concentrated at the ends of the filament in an oriented pair of positive/negative fluctuations, note that it is non-vanishing in the intermediate regions along the filament due to the varying radius of curvature of the spiral.
%\footnote{ \revisit{Why do we use this specific spiral - Its easy to compute the tangent to the spiral at any point.}}.

@@

%\revisit{A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but no $B$-mode signal is above the noise.  We predict that it should be there in higher fidelity data.}
A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but claim no $B$-mode signal above the noise. We predict that a B-mode signal from filaments should be present, since they have a finite length of a few degrees.  Detecting B-modes from filaments will be easier with more signal-to-noise and may require a more careful filament analysis, rescaling and aligning the filament ends.  While the detectability of the $B$-mode signature from stacked filaments in Planck data calls for a more careful assessment, we should see it in higher-fidelity data.

>>>>

The $B$ part of the Green's function, traced along the filament, cancels itself except at the filament ends.  This results in a non-zero $B$ pattern for the filament.  For a North--South filament, the $B$-mode pattern is positive on the North-East and South-West, and negative in the North-West and South-East.  The size of this $B$-mode pattern is set by the dimensions of the filament, chiefly the width.  In contrast to the kernel radial fuctions, which are more compact at higher bandlimit, the filament $B$-mode  pattern does not depend much on the bandlimit, provided it is sufficient to resolve the filament structure.  A filament with a more gradual edge has a $B$-pattern that is more spread out, but follows the same general structure.

The non-zero $B$ result is somewhat surprising given that the polarization pattern is symmetric to both horizontal and vertical reflections through the filament center.  However, unlike a circular ring, this filament is not a configuration with a definite parity.  Since the scalar description of polarization is coordinate independent, the $E/B$ patterns do not depend on the orientation of the filament.  For example, a filament inclined at $45^\circ$ will have a similarly inclined $E/B$ pattern, but different reflection symmetries.

Changing the polarization direction within the filament changes the $E/B$ patterns.  A $90^\circ$ rotation of the polarization with respect to the filament changes the sign of both $E$ and $B$.  In a polarization pattern aligned at $45^\circ$ to the filament, the $E$ pattern will swap with the $B$ pattern.  The work in \cite{Zaldarriaga2001a} correctly argued that linear filaments (infinite and without ends) can only produce $E$-power, but we find that for realistic filaments, the ends produce $B$-mode power with a clear signature.  Careful study of the $E/B$ mode power in filaments can provide insights into the orientation of the magnetic field with respect to the axis of the filament.

The intuition from the real-space kernels holds also when we distort the shape of the filament.  If the filament were bent around into a circle, the positive and negative parts of the $B$ pattern will cancel, and we are left with a hoop of pure $E$ pattern. Here it is important to note that this cancellation will not happen for an ellipse or any loop of non-constant radius of curvature.  The same general description holds for a spiral-shaped filament, which can can be viewed as distortion of the straight filament.  The filament is highlighted by positive $E>0$.  The $E$-pattern is more negative on the interior of a curve than on the exterior, and the concentric rings of filamentary structure make an increasingly negative $E$ value inside. While the $B$-pattern is again concentrated at the ends of the filament in an oriented pair of positive/negative fluctuations, note that it is non-vanishing in the intermediate regions along the filament due to the varying radius of curvature of the spiral.  A spiral turning with the other handedness will have a $B$-pattern with the opposite sign.
%\footnote{ \revisit{Why do we use this specific spiral - Its easy to compute the tangent to the spiral at any point.}}.

@@

%\revisit{A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but no $B$-mode signal is above the noise.  We predict that it should be there in higher fidelity data.}
A stacking analysis of Planck data \citep{2016A&A...586A.141P} sees $E>0$ along filaments (selected from intensity data), but claims no $B$-mode signal above the noise. We predict that a B-mode signal from filaments should be present, since they have a finite length of a few degrees.  Detecting B-modes from filaments will be easier with more signal-to-noise and may require a more careful filament analysis, rescaling and aligning the filament ends.  While the detectability of the $B$-mode signature from stacked filaments in Planck data calls for a more careful assessment, we should see it in higher-fidelity data.
